---
title: "Lab4"
author: "Sai Ravuru"
date: "12/11/2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## EDA

```{r}
library(car)

data = read.csv("crime.csv", header=TRUE, fill=TRUE, stringsAsFactors = FALSE)


nrow(data)
str(data)
summary(data)

head(data)



```

Based on the EDA of data variables above, the anomalies are as following:
* The probability of arrest, conviction or prison sentence cannot be higher than one.

Therefore, the anomalies above will be excluded:
```{r}
data_noamomalies <- data[which((data$prbarr < 1) & (data$prbconv < 1) & (data$prbpris < 1)), ]
head(data_noamomalies)
nrow(data_noamomalies)
```

Exploring the crime rates and their associated probabilities even further:
```{r}
hist(data_noamomalies$crmrte)
hist(data_noamomalies$prbarr)
hist(data_noamomalies$prbconv)
hist(data_noamomalies$prbpris)
```
The crimes committed per person, the probability of arrest and the probability of conviction have a very positively skewed distribution. On the other hand, the probability of prison sentence has a relatively normal distribution.

Exploring the dependent variables such as police per capita and people per square mile:
```{r}
hist(data_noamomalies$polpc)
hist(data_noamomalies$density)
```
The positively skewed distributions indicate that transformation of the variables is needed. This is due to the inverse potential relationship between the police per capita and people per square mile.


```{r}

```

